---
title: "Make a table with ligands and receptors in R with OmnipathR"
author: "Kamil Slowikowski"
date: "2020-11-24"
layout: post
tags: 
  - R
  - Tutorials
categories: notes
thumb: /notes/omnipath/ligands-receptors.png
twitter:
  card: "summary_large_image"
---

<script src="https://code.jquery.com/jquery-3.5.1.min.js" integrity="sha256-9/aliU8dGd2tb6OSsuzixeV4y/faTqgFtohetphbbj0=" crossorigin="anonymous"></script>

<link rel="stylesheet" href="//cdn.datatables.net/1.10.22/css/jquery.dataTables.min.css">

<script type="text/javascript" src="//cdn.datatables.net/1.10.22/js/jquery.dataTables.min.js"></script>


```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(
  echo = TRUE
)
options(
  markdown.HTML.header = system.file('misc', 'datatables.html', package = 'knitr')
)

library(ggplot2)
#' @importFrom ggplot2 theme_classic theme element_rect element_line
#'   element_blank element_text
#' @importFrom grid unit
theme_kamil <- theme_classic(base_family = "Helvetica") +
theme(
  panel.spacing    = unit(2, "lines"),
  panel.border     = element_rect(size = 0.5, fill = NA),
  axis.ticks       = element_line(size = 0.4),
  axis.line        = element_blank(),
  plot.title       = element_text(size = 16),
  plot.subtitle    = element_text(size = 14),
  strip.background = element_blank(),
  strip.text       = element_text(size = 16),
  legend.text      = element_text(size = 16),
  legend.title     = element_text(size = 16),
  axis.text        = element_text(size = 16),
  axis.title       = element_text(size = 16)
)

scientific_10 <- function(x) {
  ifelse(x < 0.01,
    gsub("e", "%*%10^", scales::scientific_format(digits = 1)(x)),
    signif(x, 1)
  )
}
theme_set(theme_kamil)

library(scales)

```

Curated lists of genes help computational biologists to focus analyses on a
subset of genes that might be important for a research question. For example,
we might be interested to focus on the genes encoding the signals and receptors
for cell-to-cell communication. [OmnipathR] is a new R package that provides
access to a vast database of genes called [OmniPath], organized and curated by
the [Saez-Rodriguez Lab](https://saezlab.org/). Let's try to use OmnipathR to
create a simple table with ligands and receptors.

[OmniPathR]: https://github.com/saezlab/OmnipathR
[OmniPath]: https://omnipathdb.org/

<!--more-->


# Install OmnipathR

Let's get the latest development version from GitHub. For me, that was
`r packageVersion("OmnipathR")` when I wrote this post.

```{r, eval=FALSE}
devtools::install_github("saezlab/OmnipathR")
```

```{r, include=FALSE}
dir.create("cache", showWarnings = FALSE)
```

# Make a table of ligands and receptors

Load the package.

```{r, message=FALSE, warning=FALSE}
library(OmnipathR)
library(dplyr)
```

OmniPath has a lot of annotations and
[functions](https://saezlab.github.io/OmnipathR/reference/) for accessing them.

Let's try the
[import_intercell_network()](https://saezlab.github.io/OmnipathR/reference/import_intercell_network.html)
function.

```{r, include=FALSE}
icn_file <- "cache/omnipath-icn.rds"
if (!file.exists(icn_file)) {
  icn <- OmnipathR::import_intercell_network()
  saveRDS(icn, icn_file)
} else {
  icn <- readRDS(icn_file)
}
```

```{r, eval=FALSE}
icn <- OmnipathR::import_intercell_network()
```

The returned `icn` object is a data frame with `r comma(nrow(icn))` rows and 
`r ncol(icn)` columns. It's a lot of information!

```{r include=FALSE}
write.table(icn, "omnipath-intercell-network.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
tsv_size <- utils:::format.object_size(
  file.size("omnipath-intercell-network.tsv"),
  "auto"
)
```


## Download the table

Download the entire table with `r comma(nrow(icn))` rows:

&#128190; [omnipath-intercell-network.tsv](../icn.tsv) (`r tsv_size`)


## Explore the first 100 records

Lets have a look at the first 100 records:

```{r results='asis', echo=FALSE}
kable(head(icn, 100), 'html', table.attr='id="icn_table"')
```

<script type="text/javascript">
  $(document).ready(function() {
		$('#icn_table').DataTable({
      scrollX: 400
    });
	} );
</script>

There are `r ncol(icn)` columns to explore:

```{r}
sort(colnames(icn))
```

## Show how many references support each record

There are `r comma(sum(icn$n_references == 0))` records with 0 references, and 
`r comma(sum(icn$n_references >= 1))` records with at least 1 reference.

This histogram summarizes the number of records (y-axis) with each number of
supporting references (x-axis).

```{r, message=FALSE, warning=FALSE}
library(ggplot2)

ggplot(icn) +
  geom_histogram(aes(n_references)) +
  scale_y_continuous(trans = "log10")
```

I'm not sure what the ID numbers mean, but it looks like there are some
references and databases listed here:

```{r}
icn$references[which(icn$n_references > 2)[1:5]]
```


# Filter the table

Let's filter the table to source-target pairs where the consensus score for the
source is greater than 4. Then we can take a subset of the columns, to simplify
the table.

```{r}
omni <- icn %>%
  dplyr::filter(consensus_score_intercell_source > 4) %>%
  dplyr::select(
    target_genesymbol,
    source_genesymbol,
    is_stimulation,
    consensus_score_intercell_source
  ) %>%
  unique
```

```{r}
head(omni)
```

Suppose we have a few genes of interest:

```{r}
my_genes <- c(
  "CD274", "CXCL1", "CXCL13", "CXCR3", "CXCR5"
)
```

Are the genes in this table?

Yes, and it looks like CXCR3 and CXCR5 are labeled as "target" genes:

```{r}
my_genes[my_genes %in% omni$target_genesymbol]
```

While CD274, CXCL1, CXCL13 are labeld as "source" genes:

```{r}
my_genes[my_genes %in% omni$source_genesymbol]
```

# Learn more

Please see the OmniPath website for more details: https://omnipathdb.org/


