---
title: "Working with a sparse matrix in R"
author: "Kamil Slowikowski"
date: "2020-03-11"
layout: post
tags: 
  - R
  - Tutorials
categories: notes
thumb: "/images/sparse-matrix-p.png"
twitter:
  card: "summary_large_image"
---


````{r setup, include=FALSE}
library(seriation)
library(pheatmap)
library(knitr)
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
opts_chunk$set(
  echo = TRUE
)
````

Sparse matrices are necessary for dealing with large single-cell RNA-seq
datasets. They require less memory than dense matrices, and they allow some
computations to be more efficient. In this note, we'll discuss the internals of
the `dgCMatrix` class with examples.

<!--more-->

# Install and load libraries

Let's get started by installing and loading the [Matrix] package, which
provides the sparse matrix classes that we use in this note.

```{r, eval = FALSE}
install.packages("Matrix")
library(Matrix)
```

Below, we'll explore two Matrix formats and their corresponding classes:

- The triplet format in class `dgTMatrix`
- The compressed column format in class `dgCMatrix`

```{r, eval = TRUE, include = FALSE}
library(Matrix)
```

[Matrix]: https://CRAN.R-project.org/package=Matrix

# The triplet format in dgTMatrix

`dgTMatrix` is a class from the [Matrix] R package that implements:

> general, numeric, sparse matrices in (a possibly redundant) triplet format

The format is easy to understand:

- Assume all unspecified entries in the matrix are equal to zero.
- Define the non-zero entries in triplet form `(i, j, x)` where:
  - `i` is the row number
  - `j` is the column number
  - `x` is the value

That's all there is to it. Let's make one:

````{r}
m <- Matrix(nrow = 3, ncol = 6, data = 0, sparse = TRUE)
m <- as(m, "dgTMatrix") # by default, Matrix() returns dgCMatrix
m[1,2] <- 10
m[1,3] <- 20
m[3,4] <- 30
m
```

And let's see what is inside:

```{r}
str(m)
```

The object has slots `i`, `j`, and `x`.

We can reconstruct the above sparse matrix like this:

```{r}
d <- data.frame(
  i = m@i + 1,  # m@i is 0-based, not 1-based like everything else in R
  j = m@j + 1,  # m@j is 0-based, not 1-based like everything else in R
  x = m@x
)
d
sparseMatrix(i = d$i, j = d$j, x = d$x, dims = c(3, 6))
```

We can convert a sparse matrix to a data frame like this:

```{r}
as.data.frame(summary(m))
```

Since `m@x` gives us access to the data values, we can easily transform
the values with `log2()`:

```{r}
m@x <- log2(m@x + 1)
```

# Matrix Market files use the triplet format

Many sparse [Matrix Market] files end with the file extension `.mtx`.

[Matrix Market]: https://math.nist.gov/MatrixMarket/formats.html

Write a Matrix Market file:

```{r}
writeMM(m, "matrix.mtx")
readLines("matrix.mtx")
```

Matrix Market file format:
- The first line is a comment (starts with `%%`).
- The next line says there are 3 rows, 6 columns, and 3 non-zero values.
- The next 3 lines describe the values in triplet format `(i, j, x)`.

Read a Matrix Market file:

```{r}
m <- readMM("matrix.mtx")
m
```

Convert from `dgTMatrix` to `dgCMatrix` with:

```{r}
as(m, "dgCMatrix")
```

# The compressed column format in dgCMatrix

`dgCMatrix` is a class from the [Matrix] R package that implements:

> general, numeric, sparse matrices in the (sorted) compressed sparse column
> format

This is the most common type of matrix that we will encounter when we are dealing
with scRNA-seq data.

Let's make a sparse matrix in the `dgCMatrix` format:

````{r}
library(Matrix)
m <- Matrix(nrow = 3, ncol = 6, data = 0, sparse = TRUE)
m[1,2] <- 10
m[1,3] <- 20
m[3,4] <- 30
m
```

Let's look inside:

```{r}
str(m)
```

The object has 6 slots, including `Dim`, `i`, `x`, and `p`.

`Dim` has dimensions of the matrix (3 rows, 6 columns):

```{r}
m@Dim
```

`x` has data values sorted column-wise (top to bottom, left to right):

```{r}
m@x
```

`i` has row indices for each data value. **Note:** `i` is 0-based, not 1-based
like everything else in R.

```{r}
m@i
```

What about `p`? It does *not* tell us which column each data value us in.

```{r}
m@p
```

`p` has the cumulative number of data values as we move from one column
to the next column, left to right. The first value is always 0, and the
length of `p` is one more than the number of columns.

We can compute `p` for any matrix:

```{r}
c(0, cumsum(colSums(m != 0)))
```

Since `p` is a cumulative sum, we can use [`diff()`][diff] to get the number of
non-zero entries in each column:

[diff]: https://stat.ethz.ch/R-manual/R-patched/library/base/html/diff.html

```{r}
diff(m@p)
colSums(m != 0)
```

The length of `p` is one more than the number of columns:

```{r}
length(m@p)
m@Dim[2] + 1
```

Given `p`, we can compute `j`:

```{r}
rep(1:m@Dim[2], diff(m@p))
```

Most of the time, it's easier to use `summary()` to convert a sparse matrix to
triplet `(i, j, x)` format.

```{r}
summary(m)$j
```

One more example might help to clarify how `i`, `x`, and `p` change as we
modify the matrix:

```{r}
# Add more values to the matrix
m[2,2] <- 50
m[2,3] <- 50
m[2,4] <- 50
m
str(m)
```

```{r, include = FALSE}
x <- colSums(m != 0)
```

We know that `p[1]` is always 0.

Column 1 has `r x[1]` values, so `p[2]` is `r m@p[2]`.

Column 2 has `r x[2]` values, so `p[3]` is `r m@p[2]` + `r x[2]` = `r m@p[2] + x[2]`.

Column 3 has `r x[3]` values, so `p[4]` is `r m@p[3]` + `r x[3]` = `r m@p[3] + x[3]`.

Column 4 has `r x[4]` values, so `p[5]` is `r m@p[4]` + `r x[4]` = `r m@p[4] + x[4]`.

Columns 5 and 6 have 0 values, so `p[6]` and `p[7]` are `r m@p[5]` + 0 = `r m@p[5]`.

# Sparse matrices use less memory than dense matrices

```{r, results = "hide", include = FALSE}
library(pryr)
```

```{r}
# A giant matrix
set.seed(1)
m <- sparseMatrix(
  i = sample(x = 1e4, size = 1e4),
  j = sample(x = 1e4, size = 1e4),
  x = rnorm(n = 1e4)
)
pryr::object_size(m)
pryr::object_size(as.matrix(m)) # Dense matrices require much more memory (RAM)
```

Compute the sparsity of the matrix:

```{r}
sparsity <- length(m@x) / m@Dim[1] / m@Dim[2]
sparsity
```

When writing Matrix Market files, remember to use gzip compression to save disk
space.

```{r, results = "hide"}
writeMM(m, "matrix.mtx")
```

The uncompressed file size:

```{r}
bytes_uncompressed <- file.size("matrix.mtx")
scales::number_bytes(bytes_uncompressed)
```

Compressing the file can save 50% of the disk space:

```{r}
system("gzip --keep matrix.mtx")
file.size("matrix.mtx.gz") / bytes_uncompressed
```

It takes about the same amount of time to read uncompressed or compressed
Matrix Market files:

```{r}
bench::mark(
  m <- readMM("matrix.mtx"),
  m <- readMM("matrix.mtx.gz")
)
```

# writeMMgz

Since the `writeMM()` function does not accept a connection object, this
does not work:

```r
writeMM(m, gzfile("matrix.mtx.gz")) ## This does not work :(
```

Instead, we can write our own function:

```{r}
#' @param x A sparse matrix from the Matrix package.
#' @param file A filename that ends in ".gz".
writeMMgz <- function(x, file) {
  mtype <- "real"
  if (is(x, "ngCMatrix")) {
    mtype <- "integer"
  }
  writeLines(
    c(
      sprintf("%%%%MatrixMarket matrix coordinate %s general", mtype),
      sprintf("%s %s %s", x@Dim[1], x@Dim[2], length(x@x))
    ),
    gzfile(file)
  )
  data.table::fwrite(
    x = summary(x),
    file = file,
    append = TRUE,
    sep = " ",
    row.names = FALSE,
    col.names = FALSE
  )
}
```

Confirm that it works:

```{r}
writeMMgz(m, "matrix2.mtx.gz")

all.equal(readMM("matrix.mtx.gz"), readMM("matrix2.mtx.gz"))
```

```{r, include = FALSE}
unlink("matrix.mtx")
unlink("matrix.mtx.gz")
unlink("matrix2.mtx.gz")
```

# Learn more

Find more details about the matrix formats in [this vignettes][1] from the
Matrix R package.

[1]: https://cran.r-project.org/web/packages/Matrix/vignettes/Intro2Matrix.pdf

And learn more about faster computations with sparse matrices in [this vignette][2].

[2]: https://cran.r-project.org/web/packages/Matrix/vignettes/Comparisons.pdf

